---
title: "Non-parametric models - Estimating conditional variance"
output: pdf_document
author: "Haoran Mo, Alexandra Yamaui"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sm)
source("lpr_visual.R")
source("locpolreg.R")
source("loc_lin_reg.R")
```

Having the heteroscedastic regression model $$Y = m(x) + \sigma(x)\varepsilon = m(x) + \epsilon,$$, where $E(\varepsilon) = 0, V(\varepsilon) = 1$

we want to estimate the function that represents the conditional variance $\sigma^2$ of the variable lgWeigth (log(Weight)) from the aircraft dataset, given that the explanatory variable (Yr) is equal to a value x. We will use nonparametric methods to do that. Below is the plot the response variable vs the explanatory variable.

```{r}
data(aircraft)
attach(aircraft)
lgWeight <-log(Weight)
plot(Yr,lgWeight)
```

Initially, we are going to fit a local linear regression model to obtain an estimation $\hat{m}(x)$ of every point $x$. The general idea is to build a grid of intervals ($t_i$) centered around each point $x$ and estimate a local linear regression in each interval. Doing this, we are going to try multiple sizes for the intervals and several values for the smoothing parameter $h$, which controls weight concentration around each point $x$.

To make the regression function smooth weights are assigned to each pair $(t_i,y_i)$ using a kernel function, which, in this case, is the Normal density function centered at 0, with $h$ as standard deviation.

The result is shown below, where we can see an smooth function (in red) of the estimates.
```{r}
# step 1 Fit a nonparametric regression to data (xi,yi) and save 
#   the estimated values mˆ (xi).
par(mfrow=c(2,2))
hs <- c(1,1.4,4,8.4,10)
ts <- c(30,51,70)
for (t in ts) {
  for (h in hs) {
    tg = seq(min(Yr), max(Yr), length=t)
    llr <- loc.lin.reg(x=Yr, y=lgWeight, h=h, tg=tg)
    plot(Yr,lgWeight, main = paste("h = ", paste(h, paste(", t = ", t, sep = ""), sep = "")))
    lines(tg, llr$mt, col=2, lwd=2)
  }
}
# TODO is this regression good enough?
```
Changing the smooth parameter and the size of the interval we can see that $h$ has more impact than sizds of interval, as $h$ increases the function curve becomes smoother. We will choose the combination $h = 4$ and $t = 70$

```{r}
h <- 4
tg <- seq(min(Yr), max(Yr), length=70)
```

Now we are going to transform the estimated residuals applying logarithm to the square of it $\hat\epsilon_i = (y_i - \hat{m}(x_i))^2$, which represents the variance (the square deviation) of the model. $$z_i = log\epsilon_i^2 = log((y_i - \hat{m}(x_i))^2)$$
```{r}
# step 2 Transform the estimated residuals hat.e? = y_i - llr$mt
z_i = log((lgWeight - llr$mt)^2)
```

Then we perform a nonparametric regression over $(x_i,z_i)$ to obtain the estimation of the (logarithm of the) variance $log\sigma^2(x)$.

```{r}
# step 3 Fit a nonparametric regression to data (Yr,z_i) and 
# call the estimated function qˆ(x).
llr2 <-loc.lin.reg(x=Yr, y=z_i, h=h, tg=tg)
plot(Yr,z_i)
lines(tg, llr2$mt, col=2, lwd=2)
```

Finally, we can obtain the conditional variance applying exponential to the estimation obtained before

```{r}
# step 4 Estimate sigma_2(x)
sigma_2 =  exp(llr2$mt)
```

Plotting the residuals against $x_i$ we superimposed the estimated function $\hat{\sigma}^2(x)$
```{r}
plot(Yr,((lgWeight - llr$mt)^2)) # residual square against x_i
lines(tg, sigma_2, col=3, lwd=3)

```

Here we present the draw of the function $\hat{m}(x)$ and the superimposed bands $\hat{m}(x) ± 1.96\hat{\sigma}(x)$
```{r}
sigma = sqrt(sigma_2)
plot(Yr,lgWeight)
lines(tg,llr$mt,col=3,lwd=3) # mt from step 1
lines(tg,llr$mt+1.96*sigma,col=4,lwd=1)
lines(tg,llr$mt-1.96*sigma,col=4,lwd=1)

```