---
title: "Advanced Statistical Modeling"
subtitle: "Non-parametric models - Alternative estimations of conditional variance"
author: "Haoran Mo, Alexandra Yamaui"
date: "November 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sm)
```

```{r}
load('boston.Rdata')
```

The second approach is the method proposed by Gasser, Stroka and Jennen-Steinmetz in 1986, which consists in linear interpolation of every point $xi$ with the previous and following observation, when they are ordered based n the value of $xi$ in an ascending order.

The linear interpolation for $xi$ is defined by:
$$\hat{y}_i = \frac{x_{i+1}-x_i}{x_{i+1}-x_{i-1}}y_i+\frac{x_{i}-x_{i-1}}{x_{i+1}-x_{i-1}}y_{i+1}=a_iy_{i-1}+b{y_{i+1}}$$
and the (unbiased) estimator of $\hat\sigma^2 = \frac{1}{n-2}\sum_{i=2}^{n-1}\frac{1}{a_i^2+b_i^2+1}\hat\varepsilon_i^2$

```{r}
calculate_residual_variance <- function(X,Y) {
  XY <- data.frame(X,Y)
  XY <- XY[order(XY$X),]
  rownames(XY) <- 1:nrow(XY)
  n <- length(Y)
  
  # Rice
  t2 = XY[-c(length(X)),"Y"]
  t1 = XY[-c(1),"Y"]
  rice.sigma_2 <- (sum((t2-t1)^2))/(2*(n-1))
  
  # Gasser, Sroka, and Jennen-Steinmetz
  summation <- 0
  
  for (i in 2:(n-1)) {
    xi <- XY[i, 'X']
    
    x.previous = XY[i-1,'X'] # x_i-1
    x.following = XY[i+1,'X'] # x_i+1
    y.previous = XY[i-1,'Y'] # y_i-1
    y.following = XY[i+1,'Y'] # y_i+1
    
    if (x.previous == x.following) {
      x.following <- min(XY[XY$X > xi,]$X)
      x.previous <- max(XY[XY$X < xi,]$X)
    }
    
    a_i <- (x.following - xi)/(x.following - x.previous)
    b_i <- (xi - x.previous)/(x.following - x.previous)
    
    y.hat_i <- a_i*y.previous + b_i*y.following
    
    residual.hat_i <- y.hat_i - XY[i, 'Y']
    summation <- summation + (residual.hat_i^2/(a_i^2 + b_i^2 + 1))
  }
  
  gasser.sigma2 <- summation/(n - 2)
  return(list(rice.sigma_2, gasser.sigma2))  
}

X <- boston.c$LSTAT
Y <- boston.c$RM
(calculate_residual_variance(X,Y))

loess.fit <- loess(RM~LSTAT, data = boston.c)
(residual.variance.loess <- var(loess.fit$residuals))

```