---
title: "Advanced Statistical Modeling"
subtitle: "Non-parametric models - Iteratively Re-Weighted Least Squares"
author: "Haoran Mo, Alexandra Yamaui"
date: "November 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
burn.injuries <- read.table("burn.cont.dat", header=TRUE)
```

In this task we are going to implement the Iteratively Re-Weighted Least Squares algorithm (IRWLS), which is the most frequently used method to solve the maximization problem of the log-likelihood function. This function is at the same time used to estimate the coefficients of the Logistic Regression model. Later we are going to use the glm() R function and compare the results.

We will used zero as the initial value of the coefficients $\beta_0$ (beta_0) and $\beta_1$ (beta_1) and we will build a new response variable $z$, which is a linear combination of the points $x$. The formula is presented below

$$z_i=\beta_0+\beta_1 x_i+\frac{y_i-p_i}{p_i(1-p_i)},\ i=1,...,n$$
where $y_i$ is the original response variable and $p_i$ is defined as below, which comes from the logistic function for the conditional distribution of the response variable $y$:

$$p_i=\frac{e^{\beta_0+\beta_1 x_i}}{1+e^{\beta_0+\beta_1 x_i}}$$

```{r}
IRWLS <- function(x,y) {
  n <- length(x)
  beta_0 <- 0
  beta_1 <- 0
  s <- 0
  p <- c()
  v <- c()
  z <- c()
  l <- 0
  convergence = 1
  #convergence != TRUE
  
  while (convergence > 0.0001) {  # we set 0.0001 instead of 0 due to computing cost concerned.
    for (i in 1:n) {
      p[i] <- exp(beta_0 + beta_1*x[i])/(1 + exp(beta_0 + beta_1*x[i]))
      z[i] <- beta_0 + beta_1*x[i] + ((y[i]-p[i])/p[i]*(1-p[i]))
      v[i] <- p[i]*(1-p[i])
    }
    
    lr <- lm(z ~ x, weights = v)
    
    
    beta_0_pre <- beta_0
    beta_1_pre <- beta_1
    beta_0 <- lr$coefficients[1]
    beta_1 <- lr$coefficients[2]
    
    convergence <- (abs(beta_0-beta_0_pre) + abs(beta_1-beta_1_pre))/2
    s <- s + 1
  }
  for (i in 1:n){
    l <- l+ y[i]*log(p[i]/(1-p[i]))+log(1-p[i])
  }
  AIC <- 4-2*l
  return(c(beta_0,beta_1,AIC))
}


x <- burn.injuries$e
y <- burn.injuries$superv
betas <- IRWLS(x,y)

glm.model <- glm(y~x, family = 'binomial')
summary(glm.model)




```

The results obtained with our IRWLS function and glm are similar, the values of $\beta_0$ are `r betas[1]` and `r glm.model$coefficients[1]`, respectively, and the values for $\beta_1$ are `r betas[2]` and `r glm.model$coefficients[2]`. The AICs are 528.96 and `r betas[3]`

Comparing the output of the IRWLS algorithm and glm() function we can see that from both methods the coefficients and their significance can be obtained, and after adding a little more code related computing, we can obtain Akaike information criterion (AIC) also. IRWLS is popularly used for fitting GLMs; we think that other parts of information from glm() can be obtained from parameters from IRWLS.