---
title: "Advanced Statistical Modeling"
subtitle: "Non-parametric models - Generalized nonparametric regression model"
author: "Haoran Mo, Alexandra Yamaui"
date: "December 9th, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(sm)
```

When we are working with nonparametric logistic model the regression function is approximated as $\theta(x)\approx\beta_0^t+\beta_1^t(x-t)$, $x$ being close to $t$ and where $\theta(x)=log(\frac{p(x)}{1-p(x)})$ and $p(x)=E(Y|X=x)$.

Fitting a local logistic regression involves choosing the right bandwidth, and in order to do that we are going to use the method of maximum log-likehood for new observations, which we will estimate using leave-one-out cross-validation. Below is the function of log-likelihood using $h$ as bandwith

$$l_{CV}(h) = \sum_{i=1}^nlog(\hat{Pr}_h^{(-1)}(Y=y_i|X=x_i)) = \sum_{i=1}^n(y_ilog(\frac{p_i^{(-i)}}{1-p_i^{(-i)}}) + log(1-p_i^{(-i)}))$$
where $(y_ilog(\frac{p_i^{(-i)}}{1-p_i^{(-i)}}) + log(1-p_i^{(-i)}))$ is the contribution to the log-likelihood function of  each observation.

In this exercise we are going to choose the bandwidth for a logistic regression model that has as response variable the difference in life expectancy between men and women (le.fm) in 132 countries. The data set used contains development indicators of the different countries. The variable le.fm always takes non-negative values, except for one country, so we are going to take 0 as minimum value.
```{r}
countries<-read.table(file="countries.txt",head=T,row.names=2,dec=",")
attach(countries)
le.fm.0 <- pmax(0,le.fm)
hist(le.fm.0,br=40)
```

We can see that the frequency of difference of life expectancy between men and women shows a Poisson-like shape, it is bounded at 0, has discrete values and it is a little bit skewed to the left. Hence, will use local Poisson regression to estimate this variable

First we are going to implement the function to choose the bandwith for local Poisson regression.


```{r}

# Write your own CV bandwidth choice script for the local Poisson regression.
# h.cv.sm.possion.R
#
# method can be equal to 'loglik.CV' (default) or 'prob.missclas.CV'
h.cv.sm.poisson <- function(x, y, rg.h = NULL, l.h = 10, method = prob.missclas.CV){
  cv.h <- numeric(l.h)
  if (is.null(rg.h)){
    hh <- c(h.select(x,y,method="cv"), h.select(x,y,method="aicc"))#,hcv(x,y))
    rg.h <- range(hh)
  }
  i <- 0
  gr.h <- exp( seq(log(rg.h[1]/1.1), log(rg.h[2]*1.1), l=l.h))
  for (h in gr.h){
    i <- i+1
    cv.h[i] <- method(x,y,h)
  }
  return(list(h=gr.h,cv.h=cv.h, h.cv = gr.h[which.min(cv.h)]))
}
# method 1
prob.missclas.CV <- function(x,y,h){
  n <- length(x)
  pred <- sapply(1:n, 
                 function(i,x,y,h){
                   sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
                 },   x,y,h)
  return(sum(abs(pred-y)>.5)/n)
}
# method 2
loglik.CV <- function(x,y,h){
  n <- length(x)
  pred <- sapply(1:n, 
                 function(i,x,y,h){
                   sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
                 },   x,y,h)
  return(-sum( y*log(pred/(1-pred)) + log(1-pred) ))
}
```

Now, we are going to use the previous function to choose the bandwidth and fit a model with le.fm.0 as response variable and infant mortality rate (inf.mort) as explanatory variable.

```{r}
h.list <- h.cv.sm.poisson(inf.mort,le.fm.0)#rg.h=c(4,11)

aux <- sm.poisson(inf.mort,le.fm.0,h=h.list$h.cv, col=1) # h=10
sm.regression(inf.mort,le.fm.0,h=h.list$h.cv,col=4,add=T) # h=10
aux.glm <- glm(le.fm.0 ~ inf.mort,family=poisson)
pred <- predict(aux.glm, 
                newdata=data.frame(inf.mort=aux$eval.points),
                type="response")
lines(aux$eval.points,pred,col=2)
legend('topright', c("h.cv.sm.poisson", "sm.regression", "glm"), col = c('black', 'blue', 'red'), lty = c(1, 1, 1))
```

From the plot we can see that the three models curves show the same behavious, meaning that the all three can explain the response variable.

Then we fit another model using this time life expectancy (life.exp) as explanatory variable
```{r}
h.list <- h.cv.sm.poisson(life.exp,le.fm.0)#rg.h=c(4,11)

aux <- sm.poisson(life.exp,le.fm.0,h=h.list$h.cv, col=1) # h=10
sm.regression(life.exp,le.fm.0,h=h.list$h.cv,col=4,add=T) # h=10
aux.glm <- glm(le.fm.0 ~ life.exp, family=poisson)
pred <- predict(aux.glm, 
                newdata=data.frame(life.exp=aux$eval.points),
                type="response")
lines(aux$eval.points,pred,col=2)
legend('topright', c("h.cv.sm.poisson", "sm.regression", "glm"), col = c('black', 'blue', 'red'), lty = c(1, 1, 1))
```

In this case, we can see that the parametric model fitted with a Poisson GLM could not adjust that well this time. This could be due to the fact that it is a parametric model and assumes a Poisson distribution. However, the preconditions or assumptions may not be met and therefore it has a poor fit.